Verification of Stencil Computations through Spatial Specifications

### 2016-07-06 paper.tex
added more detail to evaluation

\section{Evaluation}
\label{sec:evaluation}
We evaluated our stencil language on a corpus of scientific code. We first give an overview of the corpus itself before presenting our results.

@@

\textbf{Navier} is a smaller projecter, giving a discrete
approximation to the two-dimensional Navier-Stokes fluid equations,
based on the book of~\citet{griebel1997numerical}.

\textbf{CP} consists of the example code from the book ``Computational Physics''~\cite{nicholas2006computational} introducing numerical techniques and their application to modern physics problems such as fields, waves, statistical mechanics and quantum mechanics.

@@

We first examined how frequently stencil computations occur in our
corpus. We define a \emph{potential} statement as one whose left-hand
side is an array subscript on neighbourhood indices. These statements
are the inputs to Step 2 of the Inference process described
Section~\ref{subsec:inference}. Each potential statement for which we
eventually generate some stencil specifications for is also counted as
an \emph{actual} statement.

@@

The first two columns of Table~\ref{tab:tickAssign} shows the number
of potential and actual stencils and their rate as a percentage of the
total number of successfully parsed lines of code. There are a variety
of reasons why we might not infer an actual stencil from a potential
one: 1) \textbf{Non-subset induction variables} occurs when the
induction variables on the RHS are not a subset of the LHS. These
cases are not stencils by our definition. The degenereate case of this
is to have only constant indices on the LHS. We see lots of examples
of this in loops as accumulators e.g. computing the sum over some
other array; 2) \textbf{Derived induction variables} where the
indexing variable (\mintinline{fortran}{x}) is derived from an
induction variable \mintinline{fortran}{i} e.g. \mintinline{fortran}{x
  = len - i}; 3) \textbf{Inconsistent induction dimensions} occur when
an induction variable is used to specify more than one array dimension
on the RHS. These are common in matrix operations such as
LU-decomposition which contains assignments such as
\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.

@@

We grouped stencils together into four classes based on similar levels of complexity. These four classes account for 99.7\% of all the stencils we found in the corpus.

\textbf{All-reflexive} The vast majority (X\%) of stencils were
reflexive in every dimension of the array. These most often correspond
to applying point-wise transformations to data such as scaling
e.g. \begin{minted}{fortran}
ZZ1(I,K) = .000041868*ZZ1(I,K)
\end{minted}
Arrays with this class of stencil must be indexed by all the induction
variables used in the LHS without any offsets.

\textbf{Some-reflexive} stencils consist of only reflexive terms but
leave some dimensions unspecified. Two examples in this class are listed:
\begin{minted}{fortran}
leafpools(:,age_class)=leafpool(:,1)
RVM(J,I2) = VRFO(ivar,J)
\end{minted}
where for the first example \mintinline{fortran}{leafpools} contains an
implicit loop, and for the second example \mintinline{fortran}{J} is
the induction variable. Arrays with this class of stencil can be
indexed by a subset of the LHS induction variables but again without
any offsets.

\textbf{Single-action} stencils consist of a single \term{backward},
\term{forward}, or \term{centered} term and some number of reflexive
terms.
% /home/mrd45/src/corpus/geos-chem/GTMM/CasaRegridModule.F90:645
\begin{minted}{fortran}
do j=2,jm-ig
   do i=1,im
      != stencil readOnce, (backward(depth=1, dim=2))*(reflexive(dim=1)) :: dm, q
      al(i,j) = 0.5e+0_fp*(q(i,j-1)+q(i,j)) + r3*(dm(i,j-1) - dm(i,j))
   enddo
enddo
\end{minted}

\textbf{Single-action-irreflexive} stencils consist of a single
\term{backward}, \term{forward}, or \term{centered} irreflexive term
and some number of reflexive terms.
% /home/mrd45/src/corpus/geos-chem/GeosRad/rrtmg_lw_rtrnmc.F90:355
\begin{minted}{fortran}
do lev = nlayers, 1, -1
    ! ...
    radld = radld - radld * (atrans(lev) + &
       efclfrac(lev,igc) * (1. - atrans(lev))) + &
       gassrc + cldfmc(igc,lev) * &
       (bbdtot * atot(lev) - gassrc)
    != stencil readOnce, (forward(depth=1, dim=2, irreflexive)) :: cldfmc
    drad(lev-1) = drad(lev-1) + radld
\end{minted}
Notice that, in this last example, the read from
\mintinline{fortran}{cldfmc} flows through onto the line with the
stencil.

@@

>>>>

\section{Evaluation}
\label{sec:evaluation}
We evaluated the stencil language on a corpus of scientific code. We first give an overview of the corpus itself before presenting our results.

@@

\textbf{Navier} is a small project, giving a discrete
approximation to the two-dimensional Navier-Stokes fluid equations,
based on the book of~\citet{griebel1997numerical}.

\textbf{CP} consists of the example code from the second edition of the book ``Computational Physics''~\cite{nicholas2006computational} introducing numerical techniques and their application to modern physics problems such as fields, waves, statistical mechanics and quantum mechanics.

@@

We first examined how frequently stencil computations occur in our
corpus. We define a \emph{potential} statement as one whose left-hand
side is an array subscript on neighbourhood indices. These statements
are the inputs to Step 2 of the Inference process described
Section~\ref{subsec:inference}. Each potential statement for which we
eventually generate some stencil specifications is also counted as
an \emph{actual} statement.

@@

The first two columns of Table~\ref{tab:tickAssign} shows the number
of potential and actual stencils and their rate as a percentage of the
total number of successfully parsed lines of code. There are a variety
of reasons why we might not infer an actual stencil from a potential
one: 1) \textbf{Non-subset induction variables} occurs when the
induction variables on the RHS are not a subset of the LHS. These
cases are not stencils by our definition. The degenereate case of this
is to have only constant indices on the LHS. We see lots of examples
of this in loops as accumulators e.g. computing the sum over some
other array; 2) \textbf{Derived induction variables} where the
indexing variable (\mintinline{fortran}{x}) is derived from an
induction variable (\mintinline{fortran}{i}) as in \mintinline{fortran}{x
  = len - i}; 3) \textbf{Inconsistent induction dimensions} occur when
an induction variable is used to specify more than one array dimension
on the RHS. These are common in matrix operations such as
LU-decomposition which contains assignments such as
\mintinline{fortran}{a(l) = a(l) - a(m) * b(l, m)}.

@@

We grouped stencils together into four classes based on similar levels of complexity. These four classes account for 99.4\% of all the stencils we found in the corpus.

\textbf{All-reflexive} The vast majority (66\%) of stencils were
reflexive in every dimension of the array. These most often correspond
to applying point-wise transformations to data such as scaling
e.g. \begin{minted}{fortran}
ZZ1(I,K) = .000041868*ZZ1(I,K)
\end{minted}
Arrays with this class of stencil must be indexed by all the induction
variables used in the LHS without any offsets.

\textbf{Some-reflexive} stencils (24\%) consist of only reflexive terms but
leave some dimensions unspecified. Here are two examples:
\begin{minted}{fortran}
leafpools(:,age_class)=leafpool(:,1)
RVM(J,I2) = VRFO(ivar,J)
\end{minted}
where for the first example \mintinline{fortran}{leafpools} contains an
implicit loop, and for the second example \mintinline{fortran}{J} is
the induction variable. This stencil permits array indexing by a subset of the LHS induction variables but again without
any offsets.

\textbf{Single-action} stencils (6\%) consist of a single \term{backward},
\term{forward}, or \term{centered} region constant and some number of \term{reflexive}
region constants.
% /home/mrd45/src/corpus/geos-chem/GTMM/CasaRegridModule.F90:645
\begin{minted}{fortran}
do j=2,jm-ig
   do i=1,im
      != stencil readOnce, (backward(depth=1, dim=2))*(reflexive(dim=1)) :: dm, q
      al(i,j) = 0.5e+0_fp*(q(i,j-1)+q(i,j)) + r3*(dm(i,j-1) - dm(i,j))
   enddo
enddo
\end{minted}
This stencil requires a neighbourhood index including zero-offset in one dimension only. The remaining dimensions must either have no offset or not pertain to an induction variable.

\textbf{Single-action-irreflexive} stencils (3\%) ) consist of a single
\term{backward}, \term{forward}, or \term{centered} irreflexive term
and some number of reflexive terms.
% /home/mrd45/src/corpus/geos-chem/GeosRad/rrtmg_lw_rtrnmc.F90:355
\begin{minted}{fortran}
do lev = nlayers, 1, -1
    ! ...
    radld = radld - radld * (atrans(lev) + &
       efclfrac(lev,igc) * (1. - atrans(lev))) + &
       gassrc + cldfmc(igc,lev) * &
       (bbdtot * atot(lev) - gassrc)
    != stencil readOnce, (forward(depth=1, dim=2, irreflexive)) :: cldfmc
    drad(lev-1) = drad(lev-1) + radld
\end{minted}
Notice that, in this last example, the read from
\mintinline{fortran}{cldfmc} flows through onto the line with the
stencil. Stencils of this type required a neightbourhood index without a zero-offset in one dimension only. The remaining dimensions must either have no offset or not pertain to an induction variable.

The remaining stencil shapes (less than 1\%) consisted primarily of multiple \term{forward}, \term{backward}, or \term{centered} dimensions. We note that only 0.2\% of stencils found required use of the union (\term{+}) operator. Despite this low frequency these stencils are arguably the most important since they represent more complicated code and thus greater potential for programming errors.

@@

\anote{Any discussion left to do? - something on checking perhaps? access stencils?}
